---
title: "metasnf: Meta Clustering with Similarity Network Fusion in R"
output:
    html_document:
        toc: true
        toc_depth: 4
        theme: united
---

## Installation

```{r}
install.packages("metasnf")
```

## Package overview

```{r}
#library("metasnf")

devtools::load_all()

abcd_dl <- data_list(
    list(cort_t, "cortical_thickness", "neuroimaging", "continuous"),
    list(cort_sa, "cortical_area", "neuroimaging", "continuous"),
    list(subc_v, "subcortical_volume", "neuroimaging", "continuous"),
    list(income, "household_income", "demographics", "continuous"),
    list(pubertal, "pubertal_status", "demographics", "continuous"),
    uid = "unique_id"
)

set.seed(42)
abcd_sdf <- settings_df(
    abcd_dl,
    nrow = 100,
    max_k = 40
)

solutions_matrix <- batch_snf(data_list, settings_matrix, verbose = TRUE)
```

## Data set-up

### Formatting raw data

```{r}
library("metasnf")
dim(cort_t)
str(cort_t[1:4, 1:5])
```

```{r}
library("metasnf")
all_dfs <- list(
    anxiety,
    depress,
    cort_t,
    cort_sa,
    subc_v,
    income,
    pubertal
)
complete_uids <- get_complete_uids(all_dfs, uid = "unique_id")
head(complete_uids)
length(complete_uids)
anxiety_df <- dplyr::filter(anxiety, unique_id %in% complete_uids)
depress_df <- dplyr::filter(depress, unique_id %in% complete_uids)
cort_t_df <- dplyr::filter(cort_t, unique_id %in% complete_uids)
cort_sa_df <- dplyr::filter(cort_sa, unique_id %in% complete_uids)
subc_v_df <- dplyr::filter(subc_v, unique_id %in% complete_uids)
income_df <- dplyr::filter(income, unique_id %in% complete_uids)
pubertal_df <- dplyr::filter(pubertal, unique_id %in% complete_uids)
```

### Creating a data list 

```{r}
data_list <- generate_data_list(
    list(cort_t_df, "cortical_thickness", "neuroimaging", "continuous"),
    list(cort_sa_df, "cortical_surface_area", "neuroimaging", "continuous"),
    list(subc_v_df, "subcortical_volume", "neuroimaging", "continuous"),
    list(income_df, "household_income", "demographics", "continuous"),
    list(pubertal_df, "pubertal_status", "demographics", "continuous"),
    uid = "unique_id"
)
```

```{r}
summarize_dl(data_list)
```

### Target lists

```{r}
target_list <- generate_data_list(
    list(anxiety_df, "anxiety", "behaviour", "ordinal"),
    list(depress_df, "depressed", "behaviour", "ordinal"),
    uid = "unique_id"
)
summarize_dl(target_list)
```

## Defining hyperparameters for clustering 

### An overview of the settings matrix

```{r}
library("metasnf")
data_list <- generate_data_list(
    list(cort_t_df, "cortical_thickness", "neuroimaging", "continuous"),
    list(cort_sa_df, "cortical_surface_area", "neuroimaging", "continuous"),
    list(subc_v_df, "subcortical_volume", "neuroimaging", "continuous"),
    list(income_df, "household_income", "demographics", "continuous"),
    list(pubertal_df, "pubertal_status", "demographics", "continuous"),
    uid = "unique_id"
)
settings_matrix <- generate_settings_matrix(data_list)
head(settings_matrix)
```

### Building the settings matrix

#### Basic random population

```{r}
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 100
)
head(settings_matrix)
```

#### Specifying ranges of variation

```{r}
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 100,
    min_k = 10,
    max_k = 60,
    min_alpha = 0.3,
    max_alpha = 0.8
)
```

```{r}
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 20,
    k_values = c(10, 25, 50),
    alpha_values = c(0.4, 0.8)
)
```

#### Controlling data type dropout

```{r}
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 20,
    dropout_dist = "uniform"
)
head(settings_matrix)
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 20,
    dropout_dist = "none"
)

head(settings_matrix)
```

```{r}
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 20,
    min_removed_inputs = 3
)
```

#### Piecewise construction and manual editing

```{r}
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 50,
    k_values = 50
)
settings_matrix <- add_settings_matrix_rows(
    settings_matrix,
    nrow = 10,
    k_values = 80
)
```

## Generating cluster solutions

### Batch SNF

```{r}
set.seed(42)
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 20,
    min_k = 20,
    max_k = 50
)
```

```{r}
solutions_matrix <- batch_snf(data_list, settings_matrix)
```

```{r}
progressr::with_progress({
    solutions_matrix <- batch_snf(
        data_list,
        settings_matrix,
        processes = "max"
    )
})
```

### The solutions matrix

```{r}
colnames(solutions_matrix)[1:30]
solutions_matrix[1:5, "subject_NDAR_INVMIWOSHJN"]
```

```{r}
cluster_solutions <- get_cluster_solutions(solutions_matrix)
head(cluster_solutions)
```

## Meta clustering

### Defining and visualizing meta clusters

```{r}
solutions_matrix_aris <- calc_aris(solutions_matrix)
solutions_matrix_aris[1:3, 1:3]
dim(solutions_matrix_aris)
```

```{r}
meta_cluster_order <- get_matrix_order(solutions_matrix_aris)
meta_cluster_order
```

```{r}
ari_hm <- adjusted_rand_index_heatmap(
    solutions_matrix_aris,
    order = meta_cluster_order,
    show_row_names = TRUE,
    show_column_names = TRUE
)
save_heatmap(
   heatmap = ari_hm,
   path = "figures/ari_hm.png",
   width = 550,
   height = 500,
   res = 100
)
```

```{r}
shiny_annotator(ari_hm)
```

```{r}
split_vec <- c(2, 5, 12, 17)
ari_mc_hm <- adjusted_rand_index_heatmap(
   solutions_matrix_aris,
   order = meta_cluster_order,
   split_vector = split_vec
)
save_heatmap(
   heatmap = ari_mc_hm,
   path = "./ari_mc_hm.png",
   width = 550,
   height = 500,
   res = 100
)
```

### Identifying representative solutions

```{r}
rep_solutions <- get_representative_solutions(
    solutions_matrix_aris,
    split_vector = split_vec,
    order = meta_cluster_order,
    solutions_matrix
)
nrow(rep_solutions)
```

## Evaluating cluster separation across features

### Extending the solutions matrix

```{r}
extended_solutions_matrix <- extend_solutions(
    solutions_matrix,
    target_list = target_list
)
```

```{r}
esm_cols <- colnames(extended_solutions_matrix)
sm_cols <- colnames(solutions_matrix)
new_cols <- esm_cols[!esm_cols %in% sm_cols]
head(extended_solutions_matrix[, new_cols])
```

```{r}
extended_solutions_matrix <- extend_solutions(
    solutions_matrix,
    data_list = data_list,
    target_list = target_list
)
```

### Heatmap annotations 

```{r}
annotated_ari_hm <- adjusted_rand_index_heatmap(
    solutions_matrix_aris,
    order = meta_cluster_order,
    split_vector = split_vec,
    data = extended_solutions_matrix,
    top_hm = list(
        "Depression p-value" = "cbcl_depress_r_pval",
        "Anxiety p-value" = "cbcl_anxiety_r_pval",
        "Overall outcomes p-value" = "mean_pval"
    ),
    bottom_bar = list(
        "Number of Clusters" = "nclust"
    ),
    annotation_colours = list(
        "Depression p-value" = colour_scale(
            extended_solutions_matrix[, "cbcl_depress_r_pval"],
            min_colour = "purple",
            max_colour = "black"
        ),
        "Anxiety p-value" = colour_scale(
            extended_solutions_matrix[, "cbcl_anxiety_r_pval"],
            min_colour = "green",
            max_colour = "black"
        ),
        "Overall outcomes p-value" = colour_scale(
            extended_solutions_matrix[, "mean_pval"],
            min_colour = "lightblue",
            max_colour = "black"
        )
    )
)
save_heatmap(
    heatmap = annotated_ari_hm,
    path = "figures/annotated_ari_hm.png",
    width = 700,
    height = 500,
    res = 100
)
```

```{r}
extended_solutions_matrix2 <- extended_solutions_matrix |>
    dplyr::mutate(
        key_subjects_cluster_together = dplyr::case_when(
            subject_NDAR_INVLF3TNDUZ == subject_NDAR_INVLDQH8ATK ~ TRUE,
            TRUE ~ FALSE
        )
    )
annotated_ari_hm2 <- adjusted_rand_index_heatmap(
    solutions_matrix_aris,
    order = meta_cluster_order,
    split_vector = split_vec,
    data = extended_solutions_matrix2,
    top_hm = list(
        "Depression p-value" = "cbcl_depress_r_pval",
        "Anxiety p-value" = "cbcl_anxiety_r_pval",
        "Key Subjects Clustered Together" = "key_subjects_cluster_together"
    ),
    bottom_bar = list(
        "Number of Clusters" = "nclust"
    ),
    annotation_colours = list(
        "Depression p-value" = colour_scale(
            extended_solutions_matrix[, "cbcl_depress_r_pval"],
            min_colour = "purple",
            max_colour = "black"
        ),
        "Anxiety p-value" = colour_scale(
            extended_solutions_matrix[, "cbcl_anxiety_r_pval"],
            min_colour = "green",
            max_colour = "black"
        ),
        "Key Subjects Clustered Together" = c(
            "TRUE" = "blue",
            "FALSE" = "black"
        )
    )
)
```

### Meta cluster manhattan plots 

```{r}
extended_solutions_matrix <- extend_solutions(
    solutions_matrix,
    target_list = target_list,
    data_list = data_list,
    verbose = TRUE
)
rep_solutions <- get_representative_solutions(
    solutions_matrix_aris,
    split_vector = split_vec,
    order = meta_cluster_order,
    extended_solutions_matrix
)
mc_manhattan <- mc_manhattan_plot(
    rep_solutions,
    data_list = data_list,
    target_list = target_list,
    hide_x_labels = TRUE,
    point_size = 2,
    text_size = 12,
    threshold = 0.05,
    neg_log_pval_thresh = 5
)
ggplot2::ggsave(
    "figures/mc_manhattan.png",
    mc_manhattan,
    height = 6,
    width = 12
)
```

```{r}
rep_solutions_no_cort <- dplyr::select(
    rep_solutions,
    -dplyr::contains("mrisdp")
)
mc_manhattan2 <- mc_manhattan_plot(
    rep_solutions_no_cort,
    data_list = data_list,
    target_list = target_list,
    point_size = 4,
    threshold = 0.01,
    text_size = 12,
    domain_colours = c(
        "neuroimaging" = "cadetblue",
        "demographics" = "purple",
        "behaviour" = "firebrick"
    )
)
ggplot2::ggsave(
    "figures/mc_manhattan2.png",
    mc_manhattan2,
    height = 8,
    width = 12
)
```

### P-value heatmaps

```{r}
extended_solutions_matrix <- extend_solutions(solutions_matrix, target_list)
target_pvals <- get_pvals(extended_solutions_matrix)
head(target_pvals)
pval_hm <- pval_heatmap(target_pvals, order = meta_cluster_order)
save_heatmap(
    heatmap = pval_hm,
    path = "./pval_hm.png",
    width = 400,
    height = 500,
    res = 100
)
```

### Normalized mutual information

```{r}
feature_nmis <- batch_nmi(data_list[4:5], solutions_matrix, verbose = TRUE)
feature_nmis
```

## Identifying settings driving meta cluster formation

```{r}
sm_hm <- settings_matrix_heatmap(
   settings_matrix,
   order = meta_cluster_order
)
save_heatmap(
   heatmap = sm_hm,
   path = "./settings_matrix_heatmap_ordered.png",
   width = 400,
   height = 500,
   res = 75
)
```

## Measures of quality and stability

### Quality measures

```{r}
batch_snf_results <- batch_snf(
   data_list,
   settings_matrix,
   return_similarity_matrices = TRUE
)
solutions_matrix <- batch_snf_results[["solutions_matrix"]]
similarity_matrices <- batch_snf_results[["similarity_matrices"]]
silhouette_scores <- calculate_silhouettes(
   solutions_matrix,
   similarity_matrices
)
dunn_indices <- calculate_dunn_indices(
   solutions_matrix,
   similarity_matrices
)
db_indices <- calculate_db_indices(
   solutions_matrix,
   similarity_matrices
)
```

### Stability measures

```{r}
library("metasnf")

data_list <- generate_data_list(
    list(cort_t, "cortical_thickness", "neuroimaging", "continuous"),
    list(cort_sa, "cortical_area", "neuroimaging", "continuous"),
    list(subc_v, "subcortical_volume", "neuroimaging", "continuous"),
    list(income, "household_income", "demographics", "continuous"),
    list(pubertal, "pubertal_status", "demographics", "continuous"),
    uid = "unique_id"
)

set.seed(42)
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 4,
    max_k = 40
)

solutions_matrix <- batch_snf(data_list, settings_matrix)

set.seed(42)
data_list_subsamples <- subsample_data_list(
    data_list,
    n_subsamples = 100,
    subsample_fraction = 0.85
)

batch_subsample_results <- batch_snf_subsamples(
    data_list_subsamples,
    settings_matrix,
    verbose = TRUE
)
names(batch_subsample_results)
```

```{r}
subsample_cluster_solutions <- batch_subsample_results[["cluster_solutions"]]
pairwise_aris <- subsample_pairwise_aris(
    subsample_cluster_solutions,
    return_raw_aris = TRUE,
    verbose = TRUE
)
names(pairwise_aris)
ari_summary <- pairwise_aris[["ari_summary"]]
print(ari_summary)

raw_aris <- pairwise_aris[["raw_aris"]]
print(length(raw_aris))
raw_aris[[1]][1:5, 1:5]
```

```{r}
ComplexHeatmap::Heatmap(
    raw_aris[[1]],
    heatmap_legend_param = list(
        color_bar = "continuous",
        title = "Inter-Subsample\nARI",
        at = c(0, 0.5, 1)
    ),
    show_column_names = FALSE,
    show_row_names = FALSE
)
```

```{r}
coclustering_results <- calculate_coclustering(
    subsample_cluster_solutions,
    solutions_matrix,
    verbose = TRUE
)
names(coclustering_results)
```

```{r}
cocluster_dfs <- coclustering_results[["cocluster_dfs"]]
head(cocluster_dfs[[1]])
```

```{r}
cocluster_summary <- coclustering_results[["cocluster_summary"]]
print(cocluster_summary)
```

```{r}
cocluster_density(cocluster_dfs[[1]])
```

```{r}
cocluster_heatmap(
    cocluster_dfs[[1]],
    data_list = data_list,
    top_hm = list(
        "Income" = "household_income",
        "Pubertal Status" = "pubertal_status"
    ),
    annotation_colours = list(
        "Pubertal Status" = colour_scale(
            c(1, 4),
            min_colour = "black",
            max_colour = "purple"
        ),
        "Income" = colour_scale(
            c(0, 4),
            min_colour = "black",
            max_colour = "red"
        )
    )
)
```

## Evaluating generalizability through label propagation

```{r}
 assigned_splits <- train_test_assign(
    train_frac = 0.8,
    subjects = complete_uids
)
train_subs <- assigned_splits[["train"]]
test_subs <- assigned_splits[["test"]]
train_cort_t_df <- dplyr::filter(cort_t_df, unique_id %in% train_subs)
train_cort_sa_df <- dplyr::filter(cort_sa_df, unique_id %in% train_subs)
train_subc_v_df <- dplyr::filter(subc_v_df, unique_id %in% train_subs)
train_income_df <- dplyr::filter(income_df, unique_id %in% train_subs)
train_pubertal_df <- dplyr::filter(pubertal_df, unique_id %in% train_subs)
train_anxiety_df <- dplyr::filter(anxiety_df, unique_id %in% train_subs)
train_depress_df <- dplyr::filter(depress_df, unique_id %in% train_subs)
test_cort_t_df <- dplyr::filter(cort_t_df, unique_id %in% test_subs)
test_cort_sa_df <- dplyr::filter(cort_sa_df, unique_id %in% test_subs)
test_subc_v_df <- dplyr::filter(subc_v_df, unique_id %in% test_subs)
test_income_df <- dplyr::filter(income_df, unique_id %in% test_subs)
test_pubertal_df <- dplyr::filter(pubertal_df, unique_id %in% test_subs)
test_anxiety_df <- dplyr::filter(anxiety_df, unique_id %in% test_subs)
test_depress_df <- dplyr::filter(depress_df, unique_id %in% test_subs)
full_cort_t_df <- rbind(train_cort_t_df, test_cort_t_df) 
full_cort_sa_df <- rbind(train_cort_sa_df, test_cort_sa_df) 
full_subc_v_df <- rbind(train_subc_v_df, test_subc_v_df) 
full_income_df <- rbind(train_income_df, test_income_df) 
full_pubertal_df <- rbind(train_pubertal_df, test_pubertal_df) 
full_anxiety_df <- rbind(train_anxiety_df, test_anxiety_df) 
full_depress_df <- rbind(train_depress_df, test_depress_df) 
# A data list with just training subjects
train_data_list <- generate_data_list(
    list(train_cort_t_df, "cort_thick", "neuro", "continuous"),
    list(train_cort_sa_df, "cort_sa", "neuro", "continuous"),
    list(train_subc_v_df, "subcort_vol", "neuro", "continuous"),
    list(train_income_df, "income", "demo", "continuous"),
    list(train_pubertal_df, "pubertal_status", "demo", "continuous"),
    uid = "unique_id"
)
# A data list with training and testing subjects
full_data_list <- generate_data_list(
    list(full_cort_t_df, "cort_thickness", "neuro", "continuous"),
    list(full_cort_sa_df, "cort_surface_area", "neuro", "continuous"),
    list(full_subc_v_df, "subcort_volume", "neuro", "continuous"),
    list(full_income_df, "income", "demo", "continuous"),
    list(full_pubertal_df, "pubertal_status", "demo", "continuous"),
    uid = "unique_id"
)
# Construct the target lists
train_target_list <- generate_data_list(
    list(train_anxiety_df, "anxiety", "behav", "ordinal"),
    list(train_depress_df, "depressed", "behav", "ordinal"),
    uid = "unique_id"
)
# Find a clustering solution in your training data
set.seed(42)
settings_matrix <- generate_settings_matrix(
    train_data_list,
    nrow = 5,
    min_k = 10,
    max_k = 30
)
train_solutions_matrix <- batch_snf(
    train_data_list,
    settings_matrix
)
extended_solutions_matrix <- extend_solutions(
    train_solutions_matrix,
    train_target_list
)
```

```{r}
propagated_labels <- lp_solutions_matrix(
    extended_solutions_matrix,
    full_data_list
)
head(propagated_labels)
tail(propagated_labels)
```

## Additional plots

### Feature plots

```{r}
library("metasnf")
data_list <- generate_data_list(
    list(subc_v, "subc_v", "neuro", "continuous"),
    list(income, "income", "demo", "continuous"),
    list(fav_colour, "colour", "misc", "categorical"),
    list(pubertal, "pubert", "demo", "continuous"),
    list(anxiety, "anxiety", "behav", "ordinal"),
    list(depress, "depressed", "behav", "ordinal"),
    uid = "unique_id"
)
set.seed(42)
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 2,
    min_k = 20,
    max_k = 50
)
solutions_matrix <- batch_snf(data_list, settings_matrix)
sm_row <- solutions_matrix[1, ]
plot_list <- auto_plot(
    solutions_matrix_row = sm_row,
    data_list = data_list
)
```

```{r}
# Plot A
plot_list[["smri_vol_scs_csf"]]
# Plot B
plot_list[["colour"]]
# Plot C
plot_list[["colour"]] +
    ggplot2::labs(
        fill = "Favourite Colour",
        x = "Cluster",
        title = " Favourite Colour by Cluster"
    ) +
    ggplot2::scale_fill_manual(
        values = c(
            "green" = "forestgreen",
            "red" = "firebrick3",
            "yellow" = "darkgoldenrod1"
        )
    )
```

### Correlation plots

```{r}
library("metasnf")
cort_sa_minimal <- cort_sa[, 1:5]
data_list <- generate_data_list(
    list(cort_sa_minimal, "cortical_sa", "neuroimaging", "continuous"),
    list(income, "household_income", "demographics", "ordinal"),
    list(pubertal, "pubertal_status", "demographics", "continuous"),
    list(fav_colour, "favourite_colour", "demographics", "categorical"),
    list(anxiety, "anxiety", "behaviour", "ordinal"),
    list(depress, "depressed", "behaviour", "ordinal"),
    uid = "unique_id"
)
assoc_pval_matrix <- calc_assoc_pval_matrix(data_list)
ap_heatmap <- assoc_pval_heatmap(assoc_pval_matrix)
save_heatmap(
    ap_heatmap,
    "figures/ap_hm.png",
    width = 650,
    height = 500,
    res = 100
)
```

```{r}
ap_heatmap2 <- assoc_pval_heatmap(
    assoc_pval_matrix,
    confounders = list(
        "Colour" = "colour",
        "Pubertal Status" = "pubertal_status"
    ),
    out_of_models = list(
        "Income" = "household_income"
    ),
    data_list = data_list,
    split_by_domain = TRUE
)
save_heatmap(
    ap_heatmap2,
    "figures/ap_hm2.png",
    width = 700,
    height = 500,
    res = 100
)
```

### Alluvial plots 

```{r}
library("metasnf")
data_list <- generate_data_list(
    list(
        data = expression_df,
        name = "genes_1_and_2_exp",
        domain = "gene_expression",
        type = "continuous"
    ),
    list(
        data = methylation_df,
        name = "genes_1_and_2_meth",
        domain = "gene_methylation",
        type = "continuous"
    ),
    list(
        data = gender_df,
        name = "gender",
        domain = "demographics",
        type = "categorical"
    ),
    list(
        data = diagnosis_df,
        name = "diagnosis",
        domain = "clinical",
        type = "categorical"
    ),
    uid = "patient_id"
)
set.seed(42)
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 1,
    max_k = 40
)
batch_snf_results <- batch_snf(
    data_list,
    settings_matrix,
    return_similarity_matrices = TRUE
)
solutions_matrix <- batch_snf_results[["solutions_matrix"]]
similarity_matrices <- batch_snf_results[["similarity_matrices"]]
similarity_matrix <- similarity_matrices[[1]]
cluster_solution <- get_cluster_solutions(solutions_matrix)[["1"]]
cluster_sequence <- list(
    spectral_two,
    spectral_three,
    spectral_four
)
alluvial_cluster_plot(
    cluster_sequence = cluster_sequence,
    similarity_matrix = similarity_matrix,
    data_list = data_list,
    key_outcome = "gender", # the name of the feature of interest
    key_label = "Gender", # how the feature of interest should be displayed
    extra_outcomes = "diagnosis", # more features to plot but not colour by
    title = "Gender Across Cluster Counts"
)
```

### Similarity matrix heatmaps

```{r}
similarity_matrix_hm <- similarity_matrix_heatmap(
    similarity_matrix = similarity_matrix,
    cluster_solution = cluster_solution,
    heatmap_height = grid::unit(10, "cm"),
    heatmap_width = grid::unit(10, "cm")
)
similarity_matrix_hm
save_heatmap(
    heatmap = similarity_matrix_hm,
    path = "figures/similarity_matrix_hm.png",
    width = 410,
    height = 330,
    res = 80
)
```

### Other manhattan plots

```{r}
full_data_list <- generate_data_list(
    list(subc_v, "subcortical_volume", "neuroimaging", "continuous"),
    list(income, "household_income", "demographics", "continuous"),
    list(pubertal, "pubertal_status", "demographics", "continuous"),
    list(anxiety, "anxiety", "behaviour", "ordinal"),
    list(depress, "depressed", "behaviour", "ordinal"),
    uid = "unique_id"
)
data_list <- full_data_list[1:3]
target_list <- full_data_list[4:5]

set.seed(42)
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 20,
    min_k = 20,
    max_k = 50
)

solutions_matrix <- batch_snf(data_list, settings_matrix)

extended_solutions_matrix <- extend_solutions(
    solutions_matrix,
    data_list = data_list,
    target_list = target_list,
    min_pval = 1e-10
)

esm_manhattan <- esm_manhattan_plot(
    extended_solutions_matrix[1:5, ],
    neg_log_pval_thresh = 5,
    threshold = 0.05,
    point_size = 3,
    jitter_width = 0.1,
    jitter_height = 0.1,
    plot_title = "Feature-Solution Associations",
    text_size = 14,
    bonferroni_line = TRUE
)

var_manhattan <- var_manhattan_plot(
    data_list,
    key_var = "household_income",
    plot_title = "Correlation of Features with Household Income",
    text_size = 16,
    neg_log_pval_thresh = 3,
    threshold = 0.05
)
```

## Helper functions 

### Data list management

```{r}
summarize_dl(data_list)
```

```{r}
summarize_dl(data_list, scope = "feature")
```

```{r}
class(collapse_dl(data_list))
```

## Alternative formats for data list generation 

### Named nested components

```{r}
library("metasnf")
heart_rate_df <- data.frame(
    unique_id = c("1", "2", "3"),
    var1 = c(0.04, 0.1, 0.3),
    var2 = c(30, 2, 0.3)
)
personality_test_df <- data.frame(
    unique_id = c("1", "2", "3"),
    var3 = c(900, 1990, 373),
    var4 = c(509, 2209, 83)
)
survey_response_df <- data.frame(
    unique_id = c("1", "2", "3"),
    var5 = c(1, 3, 3),
    var6 = c(2, 3, 3)
)
city_df <- data.frame(
    unique_id = c("1", "2", "3"),
    var7 = c("toronto", "montreal", "vancouver")
)
data_list <- generate_data_list(
    list(
        data = heart_rate_df,
        name = "heart_rate",
        domain = "clinical",
        type = "continuous"
    ),
    list(
        data = personality_test_df,
        name = "personality_test",
        domain = "surveys",
        type = "continuous"
    ),
    list(
        data = survey_response_df,
        name = "survey_response",
        domain = "surveys",
        type = "ordinal"
    ),
    list(
        data = city_df,
        name = "city",
        domain = "location",
        type = "categorical"
    ),
    uid = "unique_id"
)
```

### List of lists

```{r}
list_of_lists <- list(
    list(heart_rate_df, "data1", "domain1", "continuous"),
    list(personality_test_df, "data2", "domain2", "continuous")
)
data_list <- generate_data_list(
    list_of_lists,
    uid = "unique_id"
)
```

## Clustering algorithms 

### The default clustering algorithms list

```{r}
clust_algs_list <- generate_clust_algs_list()
summarize_clust_algs_list(clust_algs_list)
```

```{r}
spectral_eigen <- function(similarity_matrix) {
    estimated_n <- estimate_nclust_given_graph(
        W = similarity_matrix,
        NUMC = 2:10
    )
    number_of_clusters <- estimated_n[["Eigen-gap best"]]
    solution <- SNFtool::spectralClustering(
        similarity_matrix,
        number_of_clusters
    )
    solution_data <- list(
        "solution" = solution,
        "nclust" = number_of_clusters
    )
    return(solution_data)
}
```

### Customizing the clustering algorithms list

#### Pre-defined algorithms

```{r}
clust_algs_list <- generate_clust_algs_list(
    "two_cluster_spectral" = spectral_two,
    "five_cluster_spectral" = spectral_five
)
summarize_clust_algs_list(clust_algs_list)
```

```{r}
clust_algs_list <- generate_clust_algs_list(
    "two_cluster_spectral" = spectral_two,
    "five_cluster_spectral" = spectral_five,
    disable_base = TRUE
)
summarize_clust_algs_list(clust_algs_list)
```

```{r}
set.seed(42)
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 10,
    max_k = 40,
    clustering_algorithms = clust_algs_list
)
```

### Using a custom clustering algorithms list

```{r}
data_list <- generate_data_list(
    list(subc_v, "subc_v", "neuro", "continuous"),
    list(income, "income", "demo", "continuous"),
    list(fav_colour, "colour", "misc", "categorical"),
    list(pubertal, "pubert", "demo", "continuous"),
    list(anxiety, "anxiety", "behav", "ordinal"),
    list(depress, "depressed", "behav", "ordinal"),
    uid = "unique_id"
)
set.seed(42)
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 2,
    min_k = 20,
    max_k = 50
)
solutions_matrix <- batch_snf(
    data_list,
    settings_matrix,
    clust_algs_list = clust_algs_list
)
```

### Non-automated clustering

```{r}
batch_snf_results <- batch_snf(
    data_list,
    settings_matrix,
    clust_algs_list = clust_algs_list,
    return_similarity_matrices = TRUE
)
solutions_matrix <- batch_snf_results[["solutions_matrix"]]
similarity_matrices <- batch_snf_results[["similarity_matrices"]]
```

## Distance metrics 

```{r}
library("metasnf")
distance_metrics_list <- generate_distance_metrics_list()
summarize_dml(distance_metrics_list)
```

### Using custom distance metrics

#### Pre-defined distance metrics functions

```{r}
my_distance_metrics <- generate_distance_metrics_list(
    continuous_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance
    ),
    discrete_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance
    )
)
summarize_dml(my_distance_metrics)
```

#### User-defined distance metrics functions

### Using a custom distance metrics list

```{r}
library(SNFtool)

data(Data1)
data(Data2)

Data1$"patient_id" <- 101:(nrow(Data1) + 100)
Data2$"patient_id" <- 101:(nrow(Data2) + 100)

data_list <- generate_data_list(
    list(Data1, "genes_1_and_2_exp", "gene_expression", "continuous"),
    list(Data2, "genes_1_and_2_meth", "gene_methylation", "continuous"),
    uid = "patient_id"
)

settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 10,
    distance_metrics_list = my_distance_metrics
)

solutions_matrix <- batch_snf(
    data_list,
    settings_matrix,
    distance_metrics_list = my_distance_metrics
)
```

## Feature weighting 

```{r}
library("metasnf")
data_list <- generate_data_list(
    list(income_df, "household_income", "demographics", "ordinal"),
    list(pubertal_df, "pubertal_status", "demographics", "continuous"),
    list(fav_colour, "favourite_colour", "demographics", "categorical"),
    list(anxiety_df, "anxiety", "behaviour", "ordinal"),
    list(depress_df, "depressed", "behaviour", "ordinal"),
    uid = "unique_id"
)
my_weights <- generate_weights_matrix(
    data_list,
    nrow = 20
)
head(my_weights)
```

```{r}
uniform_weights <- generate_weights_matrix(
    data_list,
    nrow = 5,
    fill = "uniform"
)
exp_weights <- generate_weights_matrix(
    data_list,
    nrow = 5,
    fill = "exponential"
)
```

```{r}
set.seed(42)
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 5,
    min_k = 20,
    max_k = 50
)
batch_snf(
    data_list = data_list,
    settings_matrix = settings_matrix,
    weights_matrix = exp_weights
)
```

## Association boosted similarity network fusion 

```{r}
library("metasnf")
# Build data list with held-out feature just for p-value calculations
data_list_with_held_out_f <- generate_data_list(
    list(income, "income", "demographics", "ordinal"),
    list(pubertal, "pubertal_status", "demographics", "continuous"),
    list(fav_colour, "favourite_colour", "demographics", "categorical"),
    list(anxiety, "anxiety", "behaviour", "ordinal"),
    uid = "unique_id"
)
# Calculation of all pairwise p-values
assoc_pval_matrix <- calc_assoc_pval_matrix(
    data_list_with_held_out_f
)
# Extract the relevant row
colour_pk <- assoc_pval_matrix["colour", ]
# Drop self p-value
colour_pk <- colour_pk[!names(colour_pk) == "colour"]
# Convert p-values to feature weights
feature_weights <- -log10(colour_pk)/sum(-log10(colour_pk))
# Drop held-out component from data list
data_list <- data_list_with_held_out_f
data_list["favourite_colour"] <- NULL
# Build custom distance metrics list using sew_euclidean_distance
my_distance_metrics <- generate_distance_metrics_list(
    continuous_distances = list(
         "sew_euclidean" = sew_euclidean_distance
    ),
    discrete_distances = list(
         "sew_euclidean" = sew_euclidean_distance
    ),
    ordinal_distances = list(
         "sew_euclidean" = sew_euclidean_distance
    ),
    categorical_distances = list(
         "gower" = gower_distance
    ),
    mixed_distances = list(
         "gower" = gower_distance
    ),
    keep_defaults = FALSE
)
# Build custom weights matrix using the generated feature weights
my_weights <- generate_weights_matrix(
    data_list,
    nrow = 20
)
for (row in seq_len(nrow(my_weights))) {
    my_weights[row, ] <- feature_weights
}
```

## Imputation of missing data as a meta clustering parameter 

```{r}
library("metasnf")
data_list_imp1 <- generate_data_list(
    list(subc_v, "subcortical_volume", "neuroimaging", "continuous"),
    list(income, "household_income", "demographics", "continuous"),
    list(pubertal, "pubertal_status", "demographics", "continuous"),
    list(anxiety, "anxiety", "behaviour", "ordinal"),
    list(depress, "depressed", "behaviour", "ordinal"),
    uid = "unique_id"
)
data_list_imp2 <- generate_data_list(
    list(subc_v, "subcortical_volume", "neuroimaging", "continuous"),
    list(income, "household_income", "demographics", "continuous"),
    list(pubertal, "pubertal_status", "demographics", "continuous"),
    list(anxiety, "anxiety", "behaviour", "ordinal"),
    list(depress, "depressed", "behaviour", "ordinal"),
    uid = "unique_id"
)
set.seed(42)
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 10,
    min_k = 20,
    max_k = 50
)
solutions_matrix_imp1 <- batch_snf(data_list_imp1, settings_matrix)
solutions_matrix_imp2 <- batch_snf(data_list_imp2, settings_matrix)
solutions_matrix_imp1[, "imputation"] <- 1
solutions_matrix_imp2[, "imputation"] <- 2
solutions_matrix <- rbind(solutions_matrix_imp1, solutions_matrix_imp2)
```
