---
output: github_document
bibliography: bibliography.bib
link-citations: yes
linkcolor: blue
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Meta clustering with Similarity Network Fusion

<!-- badges: start -->
<!-- badges: end -->

## Installation

**This package is currently under active development.**

We recommend installing the latest development version of the package (it has the best stuff!):

```{r eval = FALSE}
devtools::install_github("BRANCHlab/metasnf")
```

If you require a stable version of the package, please keep track of which commit you want when installing.

```{r eval = FALSE}
devtools::install_github("BRANCHlab/metasnf@8badfd4d7acaaf3b3c7af62d52b5c9324b82cc6c")
```

## Overview

*metasnf* is a package that facilitates usage of the meta clustering paradigm described in [@caruanaMeta2006] with the similarity network fusion (SNF) data integration procedure developed in [@wangSimilarity2014].

This package enables repeated iterations of SNF with distinct clustering hyperparameters and combinations of input variables by making use of a few key data structures.

The major structures are:

### The `settings_matrix`

The `settings_matrix` is a dataframe that includes all tunable clustering settings as columns.
Each row represents a distinct set of clustering settings that will be used to produce a single clustering solution by SNF and spectral clustering.

### The `data_list`

The `data_list` is a nested list that contains all preprocessed input data that is ready to be used for clustering.
Each list within a `data_list` contains (1) a dataframe that will be used for clustering, (2) the name of that dataframe, (3) the **domain** of that dataframe, and (4) the type of data (continuous, ordinal, discrete, nominal (categorical), or mixed) of the features within that dataframe.

Domains are user-specified labels that indicate which dataframes describe similar sources of data.
For example, structural MRI data and diffusion MRI data may be grouped within a neuroimaging domain.

### The `target_list`

An optional data structure that closely resembles the `data_list`, but contains data sources which are not meant to be included as direct inputs for clustering.
Data within the `target_list` is reserved for out-of-model measures for comparing completed clustering solutions.

### The `solutions_matrix`

The `solutions_matrix` is a large dataframe that appends cluster solution columns (one column per subject) to the `settings_matrix`.
It contains all the clustering results generated by the different settings included in the `settings_matrix`.

An `solutions_matrix` can be convereted to an `extended_solutions_matrix` which contains overall regression p-values for any measures stored in an `target_list`.
The `extended_solutions_matrix` is useful in pipelines where separation across one or several variables stored in the `target_list` serve as the primary metric for automating the selection of a top clustering solution.

Alternatively, a top clustering solution can be selected from a regular `solutions_matrix` using the traditional meta clustering approach of clustering cluster solutions and manually examining a few solutions from the qualitatively distinct "meta clusters" that emerge.

## Recent updates:

- [x] Changes in function names
- [x] Capability of defining custom clustering algorithms
- [x] Added flexibility during generation of the settings_matrix
- [x] Capability of calculating solution stability across patient resamplings for all cluster solutions
- [x] Capability of calculating a wide range of quality metrics for all cluster solutions
- [x] Capability of writing affinity matrices to disk
- [x] Capability of not running the clustering on every run of SNF (only makes sense when paired with the previous change)

## Work in progress:

- [ ] Change in package name
- [ ] Capability of defining custom distance metrics (including feature weights)
- [ ] Capability of defining custom SNF-level weights
- [ ] New visualization functions and improvements to existing visualization functions
- [ ] Updated vignette(s) to fully document basic and advance usage of the package

### Lower priority:

- [ ] Adapt parallel processing variant of batch_snf to reflect the above changes
- [ ] Add tests / make package CRAN ready

## Contributions

Contributions to the package are very welcome!

If you have collaborator access, please create a new branch based on main OR a fork of the repo, make your changes, and submit a pull request.
If you do not have collaborator access, please make a fork of the repo, make your changes, and submit a pull request.

You are also welcome to simply post suggested changes in the issues tab of the repo.


## References
