---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Subtyping Concussions in the ABCD dataset (metasnf)

<!-- badges: start -->
<!-- badges: end -->

A collection of useful functions for working with the ABCD dataset.

## Installation

You can install the development version of metasnf from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")

# Recommended approach (requires SSH to be set up on your machine):
# See: https://docs.github.com/en/authentication/connecting-to-github-with-ssh/generating-a-new-ssh-key-and-adding-it-to-the-ssh-agent
remotes::install_git("git@github.com:psvelayudhan/metasnf.git")

# Alternative approach using Github personal authentication token
# See: https://github.com/settings/tokens
remotes::install_github("psvelayudhan/metasnf", auth_token = "your_pat_here")

# If this repository ever becomes public:
# devtools::install_github("psvelayudhan/metasnf")
```

## Contribution

This package was developed following the instruction in the R Packages book: https://r-pkgs.org/

If you would like to contribute to this package, you can either:

1. Create a new branch with your changes and submit a pull request (preferred)
2. Clone the package and send over a copy of your changes

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(metasnf)
## basic example code - TO-DO
```
