---
title: "Setup and Preprocessing"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Setup and Preprocessing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
div.aside { background-color:#fff2e6; }
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

1. **Load the metasnf library**

```{r}

library(metasnf)

```

2. **Import data**

Imported data should be in the form of one or several wide-form (one row per patient) dataframe objects.

The `metasnf` library loads mock data into the R environment with the correct format:

- abcd_anxiety
- abcd_depress
- abcd_cort_t
- abcd_cort_sa
- abcd_subc_v
- abcd_income
- abcd_pubertal
- abcd_colour

Here is what a couple of these objects look like:

```{r}

# Patient anxiety data
colnames(abcd_anxiety)

head(abcd_anxiety)

dim(abcd_anxiety)

class(abcd_anxiety)


# Patient favourite-colour data
colnames(abcd_colour)

head(abcd_colour)

dim(abcd_colour)

```

3. **Train/test split (optional)**

We recommend splitting the data into a



The very first step is to load your data into the R environment in the format of one or several *data.frame* objects.


Using the `prop_missing_weights` function, you can quickly check what percentage of missingness is present in your data.












