---
title: "Alluvial Plots"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Alluvial Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This vignette walks through this package's available plotting functions.

First, some data setup just as was done in the previous vignettes.

```{r}
library(SNFtool)
library(metasnf)

data(Data1)
data(Data2)

gender <- c("female", "male")
diagnosis <- c("definite asthma", "possible asthma", "no asthma")
atopic_asthma <- c("Yes", "No")

c1_genders <- sample(gender, 100, replace = TRUE, prob = c(3, 1))
c1_diagnosis <- sample(diagnosis, 100, replace = TRUE, prob = c(3, 2, 1))
c1_asthma <- sample(atopic_asthma, 100, replace = TRUE, prob = c(3, 1))
c1_ages <- sample(c(12:18), 100, replace = TRUE, prob = c(1:7))

c2_genders <- sample(gender, 100, replace = TRUE, prob = c(1, 3))
c2_diagnosis <- sample(diagnosis, 100, replace = TRUE, prob = c(1, 2, 3))
c2_asthma <- sample(atopic_asthma, 100, replace = TRUE, prob = c(1, 3))
c2_ages <- sample(c(18:30), 100, replace = TRUE,)

genders <- c(c1_genders, c2_genders)
diagnoses <- c(c1_diagnosis, c2_diagnosis)
asthmas <- c(c1_asthma, c2_asthma)
ages <- c(c1_ages, c2_ages)

additional_data <- data.frame(
    gender = genders,
    diagnosis = diagnoses,
    asthma = asthmas,
    age = ages
)

set.seed(42)
random_ids <- sample(100:999)[1:200]

Data1$"patient_id" <- random_ids
Data2$"patient_id" <- random_ids
additional_data$"patient_id" <- random_ids

data_list <- generate_data_list(
    list(Data1, "genes_1_and_2_exp", "gene_expression", "continuous"),
    list(Data2, "genes_1_and_2_meth", "gene_methylation", "continuous"),
    list(additional_data, "additional_data", "miscellaneous", "mixed"),
    uid = "patient_id"
)

settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 1,
    max_k = 40,
    seed = 42
)

batch_snf_results <- batch_snf(
    data_list,
    settings_matrix,
    return_similarity_matrices = TRUE
)

solutions_matrix <- batch_snf_results$"solutions_matrix"
similarity_matrices <- batch_snf_results$"similarity_matrices"

similarity_matrix <- similarity_matrices[[1]]

cluster_solution <- get_cluster_solutions(solutions_matrix)$"1"
```

Alluvial plots can be generated to visualize how changing the number of clusters influences the distribution of patients according to one (or a few) variables of interest.

First, assemble a list clustering algorithm functions that cover the range of the number of clusters you'd like to visualize.
The example below uses `spectral_two` to `spectral_six`, which are spectral clustering functions covering 2 clusters to 6 clusters respectively.

```{r}
# Spectral clustering functions ranging from 2 to 6 clusters
cluster_sequence <- list(
    spectral_two,
    spectral_three,
    spectral_four
)
```

Then, we can either generate an alluvial plot covering our similarity matrix over these clustering algorithms for data in a `data_list`:

```{r fig.width = 7, fig.height = 5.5}
alluvial_cluster_plot(
    cluster_sequence = cluster_sequence,
    similarity_matrix = similarity_matrix,
    data_list = data_list,
    key_outcome = "gender", # the name of the variable of interest
    key_label = "Gender", # how the variable of interest should be displayed
    extra_outcomes = "diagnosis", # more variables to plot but not colour by
    title = "Gender Across Cluster Counts"
)
```

Or in an external dataframe:

```{r fig.width = 7, fig.height = 5.5}
df_outcomes = data.frame(
    subjectkey = colnames(similarity_matrix),
    wheeze_status = c(
        rep("Yes", 20),
        rep("Sometimes", 60),
        rep("No", 120)
    ),
    asthma_diagnosis = c(
        rep("Definite Asthma", 10),
        rep("Possible Asthma", 50),
        rep("No Asthma", 140)
    )
)

alluvial_cluster_plot(
    cluster_sequence = cluster_sequence,
    similarity_matrix = similarity_matrix,
    data = df_outcomes,
    key_outcome = "wheeze_status",
    key_label = "Wheeze Status",
    extra_outcomes = "asthma_diagnosis",
    title = "Wheeze Status Across Cluster Counts"
)
```
