---
title: "Unwanted Signal"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Unwanted Signal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
div.aside { background-color:#fff2e6; }
</style>

```{r, include = FALSE}
# Default chunk options
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.5,
  fig.align = "center"
)
```

Download a copy of the vignette to follow along here: [unwanted_signal.Rmd](https://raw.githubusercontent.com/BRANCHlab/metasnf/main/vignettes/unwanted_signal.Rmd)

This vignette walks through handling confounding variables and variables introducing unwanted signal using the metasnf package.

A confounding variable is one that influences both a dependent and independent variable, making it appear as though there is a stronger association between the two than there mechanistically is.

An unwanted variable (terminology which is being defined here) is a variable that a user explicitly does not want their subtyping solution to be strongly associated with.

## Confounding variables

We'll start with confounding variables - that part is easy!
Really, there's nothing extra to be done.
Simply include the confounding variable into the cluster analysis and bear in mind that the separation across clusters you see in any variable is not evidence of a causal link.

Consider the following scenario.
Children raised in wealthy homes can afford to go to better schools, are likely to be raised in less stressful environments, and can afford tutoring outside of school.
Consequently, they have higher standardized test scores.
They are also substantially more likely to live near a lake.
If you were to cluster children according to their standardized test scores and whether or not they live near a lake, you may find one cluster that has kids who live near a lake and who have high test scores and a separate cluster that has kids who don't live near a lake and have low test scores.
Of course, there is no causal relationship indicating that living near a lake improves your test scores, or that your test scores causally influence your proximity to lakes.
The confounding variable is household income.
By including household income in the model, you'll get the same clusters, but those clusters will also separate across the "confounding" variable.

## Unwanted signal

Removing unwanted signal from a clustering solution is a little bit more complicated.

Consider the following scenario.
You wish to find clusters of basketball players.
For simplicity, let's say there are three basketball positions:

* point guards, who have a lot of assists and very few blocks
* centers, who have very few assists and lots of blocks
* shooting guards, who have a moderate amount of assists and blocks (and a lot of personal fouls)

You are handed a dataset containing the assists and blocks per season of a large number of basketball players.
However, half of the basketball players are from regular players at a community center and the other half are from the NBA.

What does the data look like?

```{r}
library(metasnf)
library(SNFtool)
library(ggplot2)

# Generating the player data
###############################################################################
# Generating function for the blocks-per-season of a player
generate_blocks <- function(level, position) {
    # The average blocks per season of all basketball players
    blocks <- rnorm(n = 1, mean = 500, sd = 50)
    # Effect of playing in the NBA
    if (level == "NBA") {
        blocks <- blocks + rnorm(n = 1, mean = 2000, sd = 100)
    } else {
        # Match the noisiness of the NBA players
        blocks <- blocks + rnorm(n = 1, mean = 0, sd = 100)
    }
    # Effect of the player's position
    if (position == "pg") {
        blocks <- blocks + 0 # Just to be explicit about it
    } else if (position == "c") {
        blocks <- blocks + 500
    } else if (position == "sg") {
        blocks <- blocks + 250
    }
    return(blocks)
}
#
# Generating function for the assists-per-season of a player
generate_assists <- function(level, position) {
    # The average assists per season of all basketball players
    assists <- rnorm(n = 1, mean = 1000, sd = 10)
    # Effect of playing in the NBA
    if (level == "NBA") {
        assists <- assists + rnorm(n = 1, mean = 2500, sd = 10)
    } else {
        # Match the noisiness of the NBA players
        assists <- assists + rnorm(n = 1, mean = 0, sd = 10)
    }
    # Effect of the player's position
    if (position == "pg") {
        assists <- assists + 400 # Just to be explicit about it
    } else if (position == "c") {
        assists <- assists + 0
    } else if (position == "sg") {
        assists <- assists + 200
    }
    return(assists)
}
#
# Helper function to fill in blocks and assists for a player given their
# position and level.
generate_player_data <- function(df) {
    df$"blocks" <- df |> apply(
        MARGIN = 1,
        FUN = function(x) {
            generate_blocks(x[[1]], x[[2]])
        }
    )
    df$"assists" <- df |> apply(
        MARGIN = 1,
        FUN = function(x) {
            generate_assists(x[[1]], x[[2]])
        }
    )
    return(df)
}
#
rows <- 300
player_data <- data.frame(
    level = sample(c("regular", "NBA"), size = rows, replace = TRUE),
    position = sample(c("pg", "c", "sg"), size = rows, replace = TRUE)
) |> generate_player_data()
#
player_data$"id" <- as.character(1:nrow(player_data))
#
# Plot by position
player_data |>
    ggplot(aes(x = blocks, y = assists, shape = level, colour = position)) +
    geom_point(size = 5, alpha = 0.3) +
    theme_bw()
```

Clustering 1: manually

```{r}

set.seed(42)
manual_data <- player_data |> dplyr::select("blocks", "assists")
manual_distance <- as.matrix(dist(manual_data))
print(sum(manual_distance))
similarity_matrix <- SNFtool::affinityMatrix(manual_distance, K = 20)

print(sum(similarity_matrix))

# n_clusters <- estimateNumberOfClustersGivenGraph(similarity_matrix, NUMC=2:10)$`Eigen-gap best`
n_clusters <- estimateNumberOfClustersGivenGraph(similarity_matrix, NUMC=2:10)$`Rotation cost best`
n_clusters
clusters <- SNFtool::spectralClustering(similarity_matrix, n_clusters)
manual_data$"clusters" <- factor(clusters)
manual_data |>
    ggplot(aes(x = blocks, y = assists, colour = clusters)) +
    geom_point(size = 5, alpha = 0.3) +
    theme_bw()

```

Clustering 2: metasnf

```{r}

set.seed(42)

metasnf_data <- player_data |> dplyr::select("id", "assists", "blocks")

data_list <- generate_data_list(
    list(
        data = metasnf_data,
        name = "player_data",
        domain = "player_data",
        type = "continuous"
    ),
    uid = "id"
)

settings_matrix <- generate_settings_matrix(
    data_list = data_list,
    nrow = 1,
    possible_snf_schemes = 1,
    k_values = 20
)

solutions_matrix <- batch_snf(data_list, settings_matrix)

batch_snf_results <- batch_snf(data_list, settings_matrix, return_similarity_matrices = TRUE)

batch_similarity_matrix <- batch_snf_results$"similarity_matrices"[[1]]

estimateNumberOfClustersGivenGraph(batch_similarity_matrix, NUMC=2:10)$`Rotation cost best`

estimateNumberOfClustersGivenGraph(batch_similarity_matrix)$`Rotation cost best`

solutions_matrix$"nclust"

cluster_solutions_df <- get_cluster_solutions(solutions_matrix) |>
    dplyr::rename(
        "id" = "subjectkey",
        "cluster" = `1`
    )
cluster_solutions_df$"cluster" <- factor(cluster_solutions_df$"cluster")

available_data$"id" <- paste0("subject_", available_data$"id")
clustered_data <- dplyr::inner_join(available_data, cluster_solutions_df, by = "id")
clustered_data |>
    ggplot(aes(x = blocks, y = assists, colour = cluster)) +
    geom_point(size = 5, alpha = 0.3) +
    theme_bw()

```

Clustering 3: metasnf

```{r}

set.seed(42)

metasnf_data <- player_data |> dplyr::select("id", "assists", "blocks")

data_list <- generate_data_list(
    list(
        data = metasnf_data,
        name = "player_data",
        domain = "player_data",
        type = "continuous"
    ),
    uid = "id"
)

settings_matrix <- generate_settings_matrix(
    data_list = data_list,
    nrow = 1,
    possible_snf_schemes = 1,
    k_values = 20,
    alpha_values = 0.5
)

solutions_matrix <- batch_snf(data_list, settings_matrix)
solutions_matrix$"nclust"

cluster_solutions_df <- get_cluster_solutions(solutions_matrix) |>
    dplyr::rename(
        "id" = "subjectkey",
        "cluster" = `1`
    )
cluster_solutions_df$"cluster" <- factor(cluster_solutions_df$"cluster")

metasnf_data$"id" <- paste0("subject_", metasnf_data$"id")
clustered_data <- dplyr::inner_join(metasnf_data, cluster_solutions_df, by = "id")
clustered_data |>
    ggplot(aes(x = blocks, y = assists, colour = cluster)) +
    geom_point(size = 5, alpha = 0.3) +
    theme_bw()

```



```{r}

player_data_df_no_id$"clusters" <- factor(clusters)
player_data_df_no_id |>
    ggplot(aes(x = blocks, y = assists, colour = clusters)) +
    geom_point(size = 5, alpha = 0.3) +
    theme_bw()

clusters <- SNFtool::spectralClustering(SNFtool::affinityMatrix(distance), K = 2)
player_data_df_no_id$"clusters" <- factor(clusters)
player_data_df_no_id |>
    ggplot(aes(x = blocks, y = assists, colour = clusters)) +
    geom_point(size = 5, alpha = 0.3) +
    theme_bw()

data_list <- generate_data_list(
    list(
        data = player_data_df,
        name = "player_stats",
        domain = "player_data",
        type = "continuous"
    ),
    uid = "id"
)

settings_matrix <- generate_settings_matrix(
    data_list = data_list,
    nrow = 1,
    possible_snf_schemes = 1
)

solutions_matrix <- batch_snf(data_list, settings_matrix)
get_clusters(solutions_matrix)

player_data$"cluster" <- factor(get_clusters(solutions_matrix))

player_data |>
    ggplot(aes(x = blocks, y = assists, shape = position, colour = cluster)) +
    geom_point(size = 5, alpha = 0.3) +
    theme_bw()
```

