---
title: "Correlation Plots"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Correlation Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Download a copy of the vignette to follow along here: [correlation_plots.Rmd](https://raw.githubusercontent.com/BRANCHlab/metasnf/main/vignettes/correlation_plots.Rmd)

In this vignette, we go through how you can visualize correlations across all the variables being used in your analyses.

## Data set-up

```{r}
library(metasnf)

# We'll just use the first few columns for this demo
abcd_cort_sa_minimal <- abcd_cort_sa[, 1:5]

# And one more mock categorical variable for demonstration purposes
abcd_city <- abcd_colour
abcd_city$"city" <- sample(
    c("toronto", "montreal", "vancouver"),
    size = nrow(abcd_city),
    replace = TRUE
)
abcd_city <- abcd_city |> dplyr::select(-"colour")

# Make sure to throw in all the data you're interested in visualizing for this
# data_list, including out-of-model measures and confounding variables.
data_list <- generate_data_list(
    list(abcd_cort_sa_minimal, "cortical_sa", "neuroimaging", "continuous"),
    list(abcd_h_income, "household_income", "demographics", "ordinal"),
    list(abcd_pubertal, "pubertal_status", "demographics", "continuous"),
    list(abcd_colour, "favourite_colour", "demographics", "categorical"),
    list(abcd_city, "city", "demographics", "categorical"),
    list(abcd_anxiety, "anxiety", "behaviour", "ordinal"),
    list(abcd_depress, "depressed", "behaviour", "ordinal"),
    uid = "patient"
)

summarize_dl(data_list)

# This matrix contains all the pairwise association p-values
association_matrix <- calculate_associations(data_list)

association_matrix[1:3, 1:3]
```

## Heatmaps

Here's what a basic heatmap looks like:

```{r eval = FALSE}
heatmap <- association_heatmap(
    association_matrix
)
```

```{r echo = FALSE}
association_heatmap(
    association_matrix
)
```

Most of this data was generated randomly, but the "colour" variable is really just a categorical mapping of "cbcl_depress".
The heatmap seems to show that quite well!

You can draw attention to confounding variables and/or any out of model measures by specifying their names as shown below.

```{r eval = FALSE}
association_heatmap(
    association_matrix,
    confounders = list(
        "Colour" = "colour",
        "Pubertal Status" = "pubertal_status"
    ),
    out_of_models = list(
        "City" = "city"
    )
)
```

```{r echo = FALSE}
heatmap <- association_heatmap(
    association_matrix,
    confounders = list(
        "Colour" = "colour",
        "Pubertal Status" = "pubertal_status"
    ),
    out_of_models = list(
        "City" = "city"
    )
)
```

The ComplexHeatmap package offers functionality for splitting heatmaps into slices.
One way to do the slices is by clustering the heatmap with k-means:

```{r eval = FALSE}
association_heatmap(
    association_matrix,
    confounders = list(
        "Colour" = "colour",
        "Pubertal Status" = "pubertal_status"
    ),
    out_of_models = list(
        "City" = "city"
    ),
    row_km = 3,
    column_km = 3
)
```

```{r echo = FALSE}
heatmap <- association_heatmap(
    association_matrix,
    confounders = list(
        "Colour" = "colour",
        "Pubertal Status" = "pubertal_status"
    ),
    out_of_models = list(
        "City" = "city"
    ),
    row_km = 3,
    column_km = 3
)
```

Another way is to explicitly pass in a categorical vector that can be used to distinguish the columns (according to their order in the association_matrix) into their desired slices.
To use the domains of the variables, you can make use of the `dl_variable_summary` function.

```{r eval = FALSE}
dl_var_summary <- dl_variable_summary(data_list)

head(dl_var_summary)

association_heatmap(
    association_matrix,
    confounders = list(
        "Colour" = "colour",
        "Pubertal Status" = "pubertal_status"
    ),
    out_of_models = list(
        "City" = "city"
    ),
    split_by_domain = dl_var_summary
)
```

```{r echo = FALSE}
dl_var_summary <- dl_variable_summary(data_list)

head(dl_var_summary)

association_heatmap(
    association_matrix,
    confounders = list(
        "Colour" = "colour",
        "Pubertal Status" = "pubertal_status"
    ),
    out_of_models = list(
        "City" = "city"
    ),
    split_by_domain = dl_var_summary
)
```
