---
title: "Distance Metrics"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Distance Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
div.aside { background-color:#fff2e6; }
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The distance_metrics_list

r-package enables users to customize what distance metrics are used in the SNF pipeline.
All information about distance metrics are stored in a `distance_metrics_list` object.

By default, `batch_snf` will create its own `distance_metrics_list` by calling the `generate_distance_metrics_list` function with no additional arguments.

```{r}
library(metasnf)

distance_metrics_list <- generate_distance_metrics_list()
```

The list is a list of functions (`euclidean_distance()` and `gower_distance()`), and printing them out directly can be messy.
The `summarize_distance_metrics_list()` function prints the object with a nicer format.

```{r}
summarize_distance_metrics_list(distance_metrics_list)
```

These lists must always contain at least 1 distance metric for each of the 5 recognized types of variables: continuous, discrete, ordinal, categorical, and mixed (any combination of the previous four).

By default, continuous, discrete, and ordinal data are converted to distance matrices using simple Euclidean distance.
Categorical and mixed data are handled using Gower's formula as implemented by the cluster package (see `?cluster::daisy`).

## How the distance_metrics_list is used

To show how the distance_metrics_list is used, we'll start by extending our distance_metrics_list beyond just the default options.
r-package provides a Euclidean distance function that applies standard normalization first, `sn_euclidean_distance()` (a wrapper around `SNFtool::standardNormalization` + `stats::dist`).
Here's how we can create a custom distance_metrics_list that includes this metric for continuous and discrete variables.

```{r}
my_distance_metrics <- generate_distance_metrics_list(
    continuous_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance
    ),
    discrete_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance
    )
)

summarize_distance_metrics_list(my_distance_metrics)
```

Now, during settings_matrix generation, we can provide our distance_metrics_list to ensure our new distance metrics will be used on some of our SNF runs.

Before making the settings_matrix, we'll quickly need to setup some data (as was done in [A Simple Example](https://branchlab.github.io/metasnf/articles/a_simple_example.html)).

```{r}
library(SNFtool)

data(Data1)
data(Data2)

Data1$"patient_id" <- 101:(nrow(Data1) + 100)
Data2$"patient_id" <- 101:(nrow(Data2) + 100)

data_list <- generate_data_list(
    list(Data1, "genes_1_and_2_exp", "gene_expression", "continuous"),
    list(Data2, "genes_1_and_2_meth", "gene_methylation", "continuous"),
    uid = "patient_id"
)
```

And then the settings_matrix can be generated:

```{r}
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 10,
    distance_metrics_list = my_distance_metrics
)

# showing just the columns that are related to distances
settings_matrix |> dplyr::select(dplyr::ends_with("dist"))
```

The continuous and discrete distance metrics values randomly fluctuate between 1 and 2, where 1 means that the first metric (`euclidean_distance()`) will be used and 2 means that the second metric (`sn_euclidean_distance`) will be used.

It's important to note that the settings_matrix itself **does not store the distance metrics**, just the pointers to which position metric in the distance_metrics_list should be used for that SNF run.
Because of that, you'll need to supply the distance_metrics_list once again when calling `batch_snf()`.

```{r eval = FALSE}
solutions_matrix <- batch_snf(
    data_list,
    settings_matrix,
    distance_metrics_list = my_distance_metrics
)
```

## Removing the default distance_metrics

There are two ways to avoid using the default distance metrics if you don't want to ever use them.

The first way is to use the `keep_defaults` parameter in `generate_distance_metrics_list()`:

```{r}

no_default_metrics <- generate_distance_metrics_list(
    continuous_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance
    ),
    discrete_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance
    ),
    ordinal_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance
    ),
    categorical_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance
    ),
    mixed_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance
    ),
    keep_defaults = FALSE
)

summarize_distance_metrics_list(no_default_metrics)
```

For this option, it is necessary to provide at least one distance metric for every variable type.
A distance_metrics_list cannot have any of its variable types be completely empty (even if you have no data for that variable type in the first place).

The second way is to explicitly specify which indices you want to sample from during settings_matrix generation:

```{r}
my_distance_metrics <- generate_distance_metrics_list(
    continuous_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance,
        "some_other_metric" = sn_euclidean_distance
    ),
    discrete_distances = list(
        "standard_norm_euclidean" = sn_euclidean_distance,
        "some_other_metric" = sn_euclidean_distance
    )
)

summarize_distance_metrics_list(my_distance_metrics)

settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 10,
    distance_metrics_list = my_distance_metrics,
    continuous_distances = 1,
    discrete_distances = c(2, 3)
)

settings_matrix |> dplyr::select(dplyr::ends_with("dist"))
```

The second option can be quite useful when paired with `add_settings_matrix_rows()`, enabling you to build up distinct blocks of rows in the settings matrix that have different combinations of distance metrics.
This can save you the trouble of needing to manage several distinct distance metrics lists or manage your solution space over separate runs of `batch_snf`.

```{r}
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 10,
    distance_metrics_list = my_distance_metrics,
    continuous_distances = 1,
    discrete_distances = c(2, 3)
)

settings_matrix <- add_settings_matrix_rows(
    settings_matrix,
    nrow = 10,
    distance_metrics_list = my_distance_metrics,
    continuous_distances = 3,
    discrete_distances = 1
)

settings_matrix |> dplyr::select(dplyr::ends_with("dist"))
```

In rows 1 to 10, only continuous data will always be handled by the first continuous distance metric while discrete data will be handled by the second and third discrete distance metrics.
In rows 11 to 20, continuous data will always be handled by the third continuous distance metric while discrete data will only be handled by the first distance metric.

## Supplying weights to distance metrics

Some distance metric functions can accept weights.
Usually, weights will be applied by direct scaling of specified variables.
In some cases (e.g. categorical distance metric functions), the way in which weights are applied may be somewhat less intuitive.
The bottom of this vignette outlines the available distance metric functions grouped by whether or not they accept weights.
You can examine the documentation of those weighted functions to learn more about how the weights you provide will be used.

An important note on providing weights for a run of SNF is that the specific form of the data may not be what you expect by the time it is ready to be converted into a distance metric function.
The "individual" and "two-step" SNF schemes involve distance metrics only being applied to the input dataframes in the `data_list` as they are.
The "domain" scheme, however, concatenates data within a domain before converting that larger dataframe into a distance matrix.
Anytime you have more than one dataframe with the same domain label and you use the domain SNF scheme, all the columns associated with that domain will be in a single dataframe when the distance metric function is applied.

The first step in providing custom weights is to generate a `weights_matrix`:

```{r}
weights_matrix <- generate_weights_matrix(
    data_list
)

weights_matrix
```

By default, this function will return a dataframe containing all the columns in the `data_list` and a single row of 1s, which are the weights that will be used for a single run of SNF.

To actually use the matrix during SNF, you'll need to make sure that the number of rows in the weights matrix is the same as the number of rows in your settings matrix.

```{r}
weights_matrix <- generate_weights_matrix(
    data_list,
    nrow = nrow(settings_matrix)
)

weights_matrix[1:5, ]
```

A `weights_matrix` of all 1s (the default `weights_matrix` used when you don't supply one yourself) 'won't actually do anything to the data.
You can either replace the 1s with your own weights that you've calculated outside of the package, or use some random weights following a uniform or exponential distribution.

```{r}
weights_matrix <- generate_weights_matrix(
    data_list,
    nrow = nrow(settings_matrix),
    fill = "uniform" # or fill = "exponential"
)

weights_matrix[1:5, ]
```

The default metrics (simple Euclidean for continuous, discrete, and ordinal data and Gower's distance for categorical and mixed data) are both capable of applying weights to data before distance matrix generation.

## Custom distance metrics

The remainder of this vignette deals with supplying custom distance metrics (including custom feature weighting).
Making use of this functionality will require a good understanding of working with functions in R.

You can also supply your own custom distance metrics.
Looking at the code from one of the package-provided distance functions shows some of the essential aspects of a well-formated distance function.

```{r}
euclidean_distance
```

First, the function should accept two arguments: `df` and `weights`, and only give one output, `distance_matrix`.
The function doesn't actually need to make use of those weights if you don't want it to.

By the time your data reaches a distance metric function, it (referred to as `df`) will always:

1. have no UID column
2. have at least one variable column
3. have no missing values
4. be a data.frame (not a tibble)

The variable column names won't be altered from the values they had when they were loaded into the data_list.

For example, consider the `abcd_anxiety` raw data supplied by r-package:

```{r}
head(abcd_anxiety)
```

Here's how to make it look more like what the distance metric functions will expect to see:

```{r}
processed_abcd_anxiety <- abcd_anxiety |>
    na.omit() |> # no NAs
    dplyr::rename("subjectkey" = "patient") |> # replace the UID with 'subjectkey'
    data.frame()

head(processed_abcd_anxiety)
```

If we want to have a distance metric that calculates Euclidean distance, but also scales the resulting matrix down such that the biggest allowed distance is a 1, it would look like this:

```{r}
my_scaled_euclidean <- function(df) {
    distance_matrix <- df |>
        stats::dist(method = "euclidean") |>
        as.matrix() # make sure it's formatted as a matrix
    distance_matrix <- distance_matrix / max(distance_matrix)
    return(distance_matrix)
}
```

Error handling will be in your hands.
For example, this function will fail if the pairwise distances for all patients is 0 (a division by 0 will occur).

Once you're happy that you distance function is working as you'd like it to:

```{r}
my_scaled_euclidean(processed_abcd_anxiety)[1:5, 1:5]
```

You can load it into a custom distance_metrics_list:

```{r}
my_distance_metrics <- generate_distance_metrics_list(
    continuous_distances = list(
        "my_scaled_euclidean" = my_scaled_euclidean
    )
)
```

## Custom weighting

Custom weighting at the variable level can get a little bit tricky.
Importantly, information of the variable weights need to be **a part of** the distance metric functions themselves; weights aren't supplied to the distance metric functions as arguments.
Below are a few examples of incorporating custom weights into distance matrices.

### Example 1: Uniform random reweighting

Here, we will set up a function that first standard normalizes all the variables, then randomly reweights all the variables according to a uniform distribution.

```{r}
# Setting up some data
processed_cort_t <- abcd_cort_t |>
    na.omit() |>
    dplyr::rename("subjectkey" = "patient") |>
    data.frame(row.names = "subjectkey")

processed_cort_t[1:5, 1:5]

uniform_weight_euclidean <- function(df) {
    df <- SNFtool::standardNormalization(df)
    # generating the random weights
    number_of_weights <- ncol(df)
    random_weights <- stats::runif(number_of_weights)
    print(random_weights[1]) # not necessary - for demonstration
    # applying the weights
    weighted_data <- as.matrix(df) %*% diag(random_weights)
    # distance calculation
    distance_matrix <- weighted_data |>
        stats::dist(method = "euclidean") |>
        as.matrix()
    return(distance_matrix)
}

uniform_weight_euclidean(processed_cort_t)[1:5, 1:5]
```

The approach of formatting variable weights in a diagonal matrix and then doing matrix multiplication with the original dataframe through the `%*%` operator is very handy to know for this purpose:

```{r}
df <- data.frame(
    col1 = c(1, 3),
    col2 = c(50, 100)
)

df

weights <- c(1, 2)

as.matrix(df) %*% diag(weights)
```

### Example 2: Fixed random weights

The approach above is limited in that you won't easily be able to find out what the actual weights being used were.

Using [closures](http://adv-r.had.co.nz/Functional-programming.html#closures), we can efficiently churn out distance metric functions that we can examine the weights of out of the gates.

```{r}
fixed_uniform_weight_euclidean <- function(max_cols) {
    random_weights <- stats::runif(max_cols)
    uniform_weight_euclidean <- function(df) {
        df <- SNFtool::standardNormalization(df)
        # generating the random weights
        number_of_weights <- ncol(df)
        random_weights_trimmed <- random_weights[1:number_of_weights]
        print(random_weights_trimmed[1]) # not necessary - for demonstration
        # applying the weights
        weighted_data <- as.matrix(df) %*% diag(random_weights_trimmed)
        # distance calculation
        distance_matrix <- weighted_data |>
            stats::dist(method = "euclidean") |>
            as.matrix()
        return(distance_matrix)
    }
    return(uniform_weight_euclidean)
}
```

This function will spit out another function that contains within it a fixed string of random uniform numbers.
The number of random numbers contained is based on the `max_cols` parameter, and should be chosen such that you're certain you'll never have a dataframe with more columns than `max_cols` that needs to be evaluated.
This may be bigger than the number of columns in any individual dataframe, depending on which SNF schemes you apply.
A safe value in this case would be something bigger than the total number of variables across all your data.

```{r}
my_metric_1 <- fixed_uniform_weight_euclidean(500)
```

Both the original uniform weight function and the closure function were given a print statement to report the first weight they apply.
Watch what happens when we repeatedly access either of these functions.

```{r}
uniform_weight <- uniform_weight_euclidean(processed_cort_t)
uniform_weight <- uniform_weight_euclidean(processed_cort_t)
uniform_weight <- uniform_weight_euclidean(processed_cort_t)

uniform_weight <- my_metric_1(processed_cort_t)
uniform_weight <- my_metric_1(processed_cort_t)
uniform_weight <- my_metric_1(processed_cort_t)
```

For the first function, a new set of random numbers is generated each time the distance metric is run.
For the second function, a new set of random numbers is only generated once (during `fixed_uniform_weight_euclidean()`).
This is particularly useful for resampling purposes.

### Example 3: Specific variable weights

You can also have specific control over how each variable is weighted.
Like before, this information will need to be built into the distance metric function.
You may want to have a good understanding of how the different SNF schemes work as they will ultimately define the possible combinations of vectors that any one dataframe requiring distance matrix conversion can have.
To actually embed the custom weights in this function, you may have to get a little creative.
Here is one approach, which involves creating a dataframe that stores all the weights we're interested in working with

```{r}
subset_of_data <- processed_cort_t[1:5, 1:3]
other_data <- processed_cort_t[1:5, 40:42]

colnames(subset_of_data)
colnames(other_data)
```

Let's pretend that is the totality of the data we're working with.

```{r}
all_variable_names <- c(colnames(subset_of_data), colnames(other_data))

number_of_variables <- length(all_variable_names)

# How many sets of reweightings we want to make
number_of_weight_combinations <- 5

number_of_weights <- number_of_weight_combinations * number_of_variables

weights_df <- matrix(
    ncol = number_of_variables,
    nrow = number_of_weight_combinations,
    data = runif(number_of_weights) # you could use something specific instead
) |> data.frame()

colnames(weights_df) <- all_variable_names

weights_df
```

Those are all the weights.
Let's say our distance metric encounters some odd subset of the data:

```{r}
odd_subset <- cbind(subset_of_data[, c(1, 2)], other_data[, c(1, 2)])

odd_subset
```

This function can take in a row from `weights_df` and spit out a function that always applies that row to weight any data it encounters:

```{r}
specific_weighted_euclidean <- function(weights_row) {
    weighted_euclidean <- function(df) {
        weights <- weights_row[, colnames(df)]
        weighted_data <- as.matrix(df) %*% diag(weights)
        distance_matrix <- weighted_data |>
            stats::dist(method = "euclidean") |>
            as.matrix()
        return(distance_matrix)
    }
    return(weighted_euclidean)
}
```

In action:

```{r}
first_row_metric <- specific_weighted_euclidean(weights_df[1, ])

first_row_metric(odd_subset)
```

### Example 4: Generating a range of weighted distance metrics using a weights matrix

(Example using the weighted hamming distance will go here...)

## List of included distance metrics

### Unweighted

These metrics can be used as is.

* `euclidean_distance`
    * simple Euclidean distance
    * applies to continuous, discrete, or ordinal data
* `sn_euclidean_distance`
    * Euclidean distance following standard normalization (SNFtool package implementation)
    * applies to continuous, discrete, or ordinal data
* `gower_distance`
    * Gower's distance (cluster package implementation)
    * applies to any data

### Weighted (closure-based)

These metrics need to be embedded with weights (see Example 4).

* `weighted_hamming_distance`
* `weighted_euclidean_distance`

And that's all there is on distance metrics!
If there's a metric you'd like to see added as a prewritten option included in the package, feel free to post an issue or make a pull request on the package's GitHub.

## References

Wang, Bo, Aziz M. Mezlini, Feyyaz Demir, Marc Fiume, Zhuowen Tu, Michael Brudno, Benjamin Haibe-Kains, and Anna Goldenberg. 2014. “Similarity Network Fusion for Aggregating Data Types on a Genomic Scale.” Nature Methods 11 (3): 333–37. https://doi.org/10.1038/nmeth.2810.
