---
title: "Label Propagation"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Label Propagation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
div.aside { background-color:#fff2e6; }
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

Download a copy of the vignette to follow along here: [label_propagation.Rmd](https://raw.githubusercontent.com/BRANCHlab/metasnf/main/vignettes/label_propagation.Rmd)

In this vignette, we will walk through label propagation in the metaSNF package.
Code from this vignette is largely taken from the end of the [less simple example vignette](https://branchlab.github.io/metasnf/articles/a_less_simple_example.html).

The label propagation procedure can be used to predict cluster membership for new, unlabeled observations based on their similarity to previously labeled observations.
These unlabeled observations could be a held out test set from your original sample or a new sample entirely.

The process involves the following steps:

1. Assign clusters to some group of patients
2. Calculate all the pairwise similarities amongst all the already clustered and to-be-labeled patients
3. Run the label propagation algorithm to predict cluster membership in the to-be-labeled patients

There is a lot of room for flexibility in how steps 1 and 2 are conducted.
SNF is not necessary at any part of the process.
For example, step one could be done by assigning clusters in your training set manually or by a simple clustering method like k-means.
Step two could be done just by calculating the euclidean distances across all the training and testing subjects for a small subset of variables.
The variables used to calculate the similarities in step 2 don't necessarily need to be the same ones used to derive the cluster solution in the training set either.

All that aside, we show here a simple approach that involves assigning the clusters by a call to `batch_snf`, assembling a data list that has the training and testing set subjects, and feeding the results into a simple label propagating function, `lp_solutions_matrix`.

```{r}
library(metasnf)

# Get a vector of all the subjects
all_subjects <- abcd_cort_t$"patient"

# Dataframe assigning 80% of subjects to train and 20% to test
train_test_split <- train_test_assign(train_frac = 0.8, subjects = all_subjects)

# Pulling the training and testing subjects specifically
train_subs <- train_test_split$"train"
test_subs <- train_test_split$"test"

# Partition a training set
train_abcd_cort_t <- abcd_cort_t[abcd_cort_t$"patient" %in% train_subs, ]
train_abcd_cort_sa <- abcd_cort_sa[abcd_cort_sa$"patient" %in% train_subs, ]
train_abcd_subc_v <- abcd_subc_v[abcd_subc_v$"patient" %in% train_subs, ]
train_abcd_h_income <- abcd_h_income[abcd_h_income$"patient" %in% train_subs, ]
train_abcd_pubertal <- abcd_pubertal[abcd_pubertal$"patient" %in% train_subs, ]
train_abcd_anxiety <- abcd_anxiety[abcd_anxiety$"patient" %in% train_subs, ]
train_abcd_depress <- abcd_depress[abcd_depress$"patient" %in% train_subs, ]

# Partition a test set
test_abcd_cort_t <- abcd_cort_t[abcd_cort_t$"patient" %in% test_subs, ]
test_abcd_cort_sa <- abcd_cort_sa[abcd_cort_sa$"patient" %in% test_subs, ]
test_abcd_subc_v <- abcd_subc_v[abcd_subc_v$"patient" %in% test_subs, ]
test_abcd_h_income <- abcd_h_income[abcd_h_income$"patient" %in% test_subs, ]
test_abcd_pubertal <- abcd_pubertal[abcd_pubertal$"patient" %in% test_subs, ]
test_abcd_anxiety <- abcd_anxiety[abcd_anxiety$"patient" %in% test_subs, ]
test_abcd_depress <- abcd_depress[abcd_depress$"patient" %in% test_subs, ]

# Find cluster solutions in the training set
train_data_list <- generate_data_list(
    list(train_abcd_cort_t, "cortical_thickness", "neuroimaging", "continuous"),
    list(train_abcd_cort_sa, "cortical_sa", "neuroimaging", "continuous"),
    list(train_abcd_subc_v, "subcortical_volume", "neuroimaging", "continuous"),
    list(train_abcd_h_income, "household_income", "demographics", "continuous"),
    list(train_abcd_pubertal, "pubertal_status", "demographics", "continuous"),
    uid = "patient"
)

settings_matrix <- generate_settings_matrix(
    train_data_list,
    nrow = 5,
    seed = 42,
    min_k = 10,
    max_k = 30
)

train_solutions_matrix <- batch_snf(
    train_data_list,
    settings_matrix
)

# Pick a solution that has good separation over our target variables
train_target_list <- generate_target_list(
    list(train_abcd_anxiety, "anxiety", "ordinal"),
    list(train_abcd_depress, "depressed", "ordinal"),
    uid = "patient"
)

extended_solutions_matrix <- extend_solutions(
    train_solutions_matrix,
    train_target_list
)
lowest_min_pval <- min(extended_solutions_matrix$"min_pval")
which(extended_solutions_matrix$"min_pval" == lowest_min_pval)
top_row <- extended_solutions_matrix[4, ]

# Propagate that solution to the subjects in the test set
# data list below has both training and testing subjects
full_data_list <- generate_data_list(
    list(abcd_cort_t, "cortical_thickness", "neuroimaging", "continuous"),
    list(abcd_cort_sa, "cortical_surface_area", "neuroimaging", "continuous"),
    list(abcd_subc_v, "subcortical_volume", "neuroimaging", "continuous"),
    list(abcd_h_income, "household_income", "demographics", "continuous"),
    list(abcd_pubertal, "pubertal_status", "demographics", "continuous"),
    uid = "patient"
)

# Use the solutions matrix from the training subjects and the data list from
# the training and testing subjects to propagate labels to the test subjects
propagated_labels <- lp_solutions_matrix(top_row, full_data_list)

head(propagated_labels)
tail(propagated_labels)
```

You could, if you wanted, see how *all* of your clustering solutions propagate to the test set, but that would mean reusing your test set and removing much of the protection against overfitting provided by this procedure.

```{r}
propagated_labels_all <- lp_solutions_matrix(
    extended_solutions_matrix,
    full_data_list
)

head(propagated_labels_all)
tail(propagated_labels_all)
```
