---
title: "Plotting Similarity Matrices"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Plotting Similarity Matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This vignette walks through usage of `similarity_matrix_heatmap` to visualize the final similarity matrix produced by a run of SNF and how that matrix associates with other patient attributes.

## Data set-up

Mock data setup is below.
This example currently makes use of the mock "Data1" and "Data2" objects that are made available by loading the SNFtool package.

```{r}
library(SNFtool)
library(metasnf)

# Load in data from SNFtool
data(Data1)
data(Data2)
```

By default, these two objects are sorted such that the first 100 patients were drawn from one distribution and the second set of 100 patients were drawn from another.
We'll extend this data with some fake additional associations.
The first 100 patients (who come from one cluster) will be more likely to be females with a higher chance of having a definite positive asthma diagnosis and of having atopic asthma than the second set of 100 patients.

```{r}

# Set the possible values
gender <- c("female", "male")
diagnosis <- c("definite asthma", "possible asthma", "no asthma")
atopic_asthma <- c("Yes", "No")

# C1: mostly F, mostly "definite asthma", mostly "Yes" to atopic asthma
c1_genders <- sample(gender, 100, replace = TRUE, prob = c(3, 1))
c1_diagnosis <- sample(diagnosis, 100, replace = TRUE, prob = c(3, 2, 1))
c1_asthma <- sample(atopic_asthma, 100, replace = TRUE, prob = c(3, 1))
c1_ages <- sample(c(12:18), 100, replace = TRUE, prob = c(1:7))

c2_genders <- sample(gender, 100, replace = TRUE, prob = c(1, 3))
c2_diagnosis <- sample(diagnosis, 100, replace = TRUE, prob = c(1, 2, 3))
c2_asthma <- sample(atopic_asthma, 100, replace = TRUE, prob = c(1, 3))
c2_ages <- sample(c(18:30), 100, replace = TRUE,)

genders <- c(c1_genders, c2_genders)
diagnoses <- c(c1_diagnosis, c2_diagnosis)
asthmas <- c(c1_asthma, c2_asthma)
ages <- c(c1_ages, c2_ages)

additional_data <- data.frame(
    gender = genders,
    diagnosis = diagnoses,
    asthma = asthmas,
    age = ages
)

head(additional_data)
```

The patient unique IDs are randomly shuffled in this example to replicate scenarios where the original data doesn't begin ordered by the true cluster labels.

```{r}

# Assign unique patient identifiers to each patient
set.seed(42)
random_ids <- sample(100:999)[1:200]

Data1$"patient_id" <- random_ids
Data2$"patient_id" <- random_ids
additional_data$"patient_id" <- random_ids

```

Then the subtyping and clustering proceeds as shown in previous vignettes.

```{r}

# Generate data_list
data_list <- generate_data_list(
    list(Data1, "genes_1_and_2_exp", "gene_expression", "continuous"),
    list(Data2, "genes_1_and_2_meth", "gene_methylation", "continuous"),
    list(additional_data, "additional_data", "miscellaneous", "mixed"),
    uid = "patient_id"
)

# Generate settings_matrix
settings_matrix <- generate_settings_matrix(
    data_list,
    nrow = 1,
    max_k = 40,
    seed = 42
)

# Run SNF and clustering
batch_snf_results <- batch_snf(
    data_list,
    settings_matrix,
    return_similarity_matrices = TRUE
)

solutions_matrix <- batch_snf_results$"solutions_matrix"
similarity_matrices <- batch_snf_results$"similarity_matrices"

# The first (and only) similarity matrix:
similarity_matrix <- similarity_matrices[[1]]

# The first (and only) cluster solution:
cluster_solution <- get_cluster_solutions(solutions_matrix)$"1"

```

## Visualize similarity matrices sorted by cluster label

`similarity_matrix_heatmap` is a wrapper for `ComplexHeatmap::Heatmap`, but with some convenient default transformations and parameters for viewing a similarity matrix.

```{r fig.width = 5, fig.height = 4.5}

similarity_matrix_heatmap(
    similarity_matrix = similarity_matrix,
    cluster_solution = cluster_solution,
    heatmap_height = grid::unit(10, "cm"),
    heatmap_width = grid::unit(10, "cm")
)

```

The default transformations include plotting log(Similarity) rather than the default similarity matrix as well as rescaling the diagonal of the matrix to the average value of the off-diagonals.
Additionally, the similarity matrix gets reordered according to the provided cluster solution.
Here's what the same similarity matrix looks like without those transformations:

```{r fig.width = 5, fig.height = 4.5}

similarity_matrix_heatmap(
    similarity_matrix = similarity_matrix,
    scale_diag = "none",
    log_graph = FALSE,
    heatmap_height = grid::unit(10, "cm"),
    heatmap_width = grid::unit(10, "cm")
)

```

## Exporting heatmaps

Here's an example of how you can save this heatmap (or any other heatmap created by `ComplexHeatmap::Heatmap`).

```{r eval = FALSE}

grDevices::png(
    "my_heatmap.png",
    width = 300,
    height = 300,
    units = "px",
)

similarity_matrix_heatmap(
    similarity_matrix = similarity_matrix,
    cluster_solution = cluster_solution
)

grDevices::dev.off()

```

## Annotations

One piece of functionality provided by `ComplexHeatmap::Heatmap` is the ability to supply visual annotations along the rows and columns of a heatmap.

You can always build annotations using the standard approaches outline in the [ComplexHeatmap Complete Reference](https://jokergoo.github.io/ComplexHeatmap-reference/book/index.html).
In addition to that, this package offers some convenient functionality to specify regular heatmap annotations and barplot annotations directly through a provided dataframe or data_list (or both).

In the example below, we make use of data supplied through a data_list.

```{r fig.width = 7, fig.height = 5.5}

# Providing the data for the annotations through a data_list
set.seed(42)
similarity_matrix_heatmap(
    similarity_matrix = similarity_matrix,
    cluster_solution = cluster_solution,
    scale_diag = "mean",
    log_graph = TRUE,
    data_list = data_list,
    left_hm = list(
        "Diagnosis" = "diagnosis"
    ),
    top_hm = list(
        "Gender" = "gender"
    ),
    top_bar = list(
        "Age" = "age"
    ),
    heatmap_height = grid::unit(10, "cm"),
    heatmap_width = grid::unit(10, "cm")
)

```

The chunk below gives an identical plot, but using a dataframe instead of a data_list.

```{r eval = FALSE}

df <- collapse_dl(data_list) # Pools all the data in a data_list to a single df

# Providing the data for the annotations through an external dataframe
set.seed(42)
similarity_matrix_heatmap(
    similarity_matrix = similarity_matrix,
    cluster_solution = cluster_solution,
    scale_diag = "mean",
    log_graph = TRUE,
    data = df,
    left_hm = list(
        "Diagnosis" = "diagnosis"
    ),
    top_hm = list(
        "Gender" = "gender"
    ),
    top_bar = list(
        "Age" = "age"
    )
)

```

We needed to call `set.seed()` because by default, `ComplexHeatmap::Heatmap` randomly generates colours for heatmap annotations.
You can instead specify your own colours using the `annotation_colours` parameter:

```{r fig.width = 7, fig.height = 5.5}

similarity_matrix_heatmap(
    similarity_matrix = similarity_matrix,
    cluster_solution = cluster_solution,
    scale_diag = "mean",
    log_graph = TRUE,
    data_list = data_list,
    left_hm = list(
        "Diagnosis" = "diagnosis"
    ),
    top_hm = list(
        "Gender" = "gender"
    ),
    top_bar = list(
        "Age" = "age"
    ),
    annotation_colours = list(
        Diagnosis = c(
            "definite asthma" = "red3",
            "possible asthma" = "pink1",
            "no asthma" = "bisque1"
        ),
        Gender = c(
            "female" = "purple",
            "male" = "lightgreen"
        )
    ),
    heatmap_height = grid::unit(10, "cm"),
    heatmap_width = grid::unit(10, "cm")
)

```

The colours `red3`, `pink1`, etc. are built-in R colours that you can browse by calling `colours()`.

For reference, the code below shows how you would achieve these annotations using standard `ComplexHeatmap` syntax.

```{r eval = FALSE}

merged_df <- collapse_dl(data_list)
order <- sort(cluster_solution, index.return = TRUE)$"ix"
merged_df <- merged_df[order, ]

top_annotations <- ComplexHeatmap::HeatmapAnnotation(
    Age = ComplexHeatmap::anno_barplot(merged_df$"age"),
    Gender = merged_df$"gender",
    col = list(
        Gender = c(
            "female" = "purple",
            "male" = "lightgreen"
        )
    ),
    show_legend = TRUE
)

left_annotations <- ComplexHeatmap::rowAnnotation(
    Diagnosis = merged_df$"diagnosis",
    col = list(
        Diagnosis = c(
            "definite asthma" = "red3",
            "possible asthma" = "pink1",
            "no asthma" = "bisque1"
        )
    ),
    show_legend = TRUE
)

similarity_matrix_heatmap(
    similarity_matrix = similarity_matrix,
    cluster_solution = cluster_solution,
    scale_diag = "mean",
    log_graph = TRUE,
    data = df,
    top_annotation = top_annotations,
    left_annotation = left_annotations
)

```

Take a look at the [ComplexHeatmap Complete Reference](https://jokergoo.github.io/ComplexHeatmap-reference/book/index.html) to learn more about what is possible with this package.

## More on sorting

Be aware that the ordering of both your data and your similarity matrix will be influenced if you supply values for the `cluster_solution` or `order` parameters.
If you don't think your data is lining up properly, consider manually making sure your `similarity_matrix` rows and columns are sorted to your preference (e.g., based on cluster) and that the order of your data matches.
This will be easier to do with a `dataframe` than with a `data_list`, as the `data_list` forces patients to be sorted by their unique IDs upon generation.
